---
title: "Should site be a covariate in the analysis model? (a9)"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
geometry: "left=3cm,right=3in,top=2cm,bottom=2cm"
output:
  pdf_document:
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
header-includes:
   - \usepackage{xcolor, graphicx, verbatim, fancyvrb, colortbl,
     longtable, caption, float}
   - \usepackage{lipsum, xargs}
   - \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
   - \usepackage[export]{adjustbox}
bibliography: /Users/zenn/p/library.bib
latex_engine: xelatex
---

```{r setup, include=FALSE}
options(dplyr.print_max = 1e9)
library(pacman)
p_load(tibble, broom, snakecase, tidyr, data.table,colorout, conflicted,
       purrr,ztable, hrbrthemes,readr, dplyr, magrittr, pipecleaner,
       lubridate, zoo, mi, survival, ggplot2, printr,
       lsmeans, ggthemes,  knitr, xtable, doBy, nlme,
       reshape2,  pander, pwr, tufte)

opts_chunk$set(warning = F, message = F, echo=F, fig.width=3.2,
	       fig.height=3,  results='asis', dev='pdf',
	       opts_chunk$set(cache.path = "cache/"))
source('~/projects/reports/templates/zz.tools.R')
options(scipen = 1, digits = 2)
options(ztable.type="latex")
```

# Introduction
\lipsum[2]
\todo[inline, color=cyan]{The original todo note withouth changed colours.\newline Here's another line.}
\lipsum[2]

Sample reference section reference. @Pocock1975a

# Methods

# Results
My most common usage of the todonotes package, is to insert an uncustomized
todonotes somewhere in a latex document. An example of this usage is the command
\todo{Make a cake \ldots},
\lipsum[1]
\todo[color=green]{Make a pie \ldots},
\lipsum[1] 
\todo{Make a joke \ldots},
\listoftodos

# References


# Introduction

A key question in the analysis of data sets from multi-site randomized clinical trials is 
whether to include site as a covariate in the model. 

Associated questions are: 

* if included, should site by treatment interaction be included? 

* should site be included as a fixed or a random effect?

* is site as a stratification factor relevant to the decision? 

* what is the experience in NEJM, JAMA, and Lancet over the last 5 years?

A review of the literature follows: 

From the clinical trials literature 

* Pros of including is explaining variance

* Con: a significant site by treatment interaction requires explaination

* Con: Including site or site x treatment costs in terms of degrees of freedom. this means 
impact on mean square error (?) (what is the
relative cost of site (say 30 sites) as a  fixed or random effect in
terms of degrees of freedom. Are there other differences, say, in
changes to other parameters in the model?)

what impact does heterogeniety of site size (num of subjects) have on analysis? 


reference @Kahan2013.  

and @Localio2001a


Given X important predictors, what is the likelihood of imbalance
if Y (<X) factors are included the stratifcation.


# Methods

```{r}
#the following function can be found at: https://stats.idre.ucla.edu/r/examples/alda/r-applied-longitudinal-data-analysis-ch-7/
#
#corandcov  function(glsob,cov=T,...){
#  corm  corMatrix(glsob$modelStruct$corStruct)[[5]]
#  print(corm)
#  varstruct  print(glsob$modelStruct$varStruct)  
#  varests  coef(varstruct, uncons=F, allCoef=T)
#  covm  corm*glsob$sigma^2*t(t(varests))%*%t(varests)
#  return(covm)
#  }
#
foo = read.table("https://stats.idre.ucla.edu/stat/r/examples/alda/data/opposites_pp.txt",header=TRUE,sep=",")
```

# Results

# References


