---
title: "Should site be a covariate in the analysis model?"
author: "R.G. Thomas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
header-includes:
  - \usepackage{xcolor, graphicx, verbatim, fancyvrb, colortbl,
    longtable, caption, float}
  - \usepackage{lipsum, xargs}
  - \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
  - \usepackage[export]{adjustbox}
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

```{r setup, include=FALSE}
options(dplyr.print_max = 1e9)
library(tibble)
library(broom)
library(tidyr)
library(purrr)
library(readr)
library(dplyr)
library(lubridate)
library(survival)
library(ggplot2)
library(knitr)
library(nlme)
library(kableExtra)
library(lme4)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 5.5, fig.height = 4,
  results = "asis", dev = "pdf",
  cache.path = "cache/"
)
options(scipen = 1, digits = 3)

source("../scripts/sim_site_covariate.R")
```

# Introduction

A key question in the analysis of data from multi-site randomized
clinical trials is whether and how to include site (or centre) as
a covariate in the analysis model. The question is not merely
academic: the choice of whether to adjust for site, and whether
to treat it as a fixed or random effect, can influence estimates
of treatment effect, standard errors, confidence interval width,
and statistical power
[@Localio2001adjustments; @Kahan2013continuous].

Multicenter trials are conducted to increase enrollment rates
and improve the generalizability of findings across clinical
settings. However, the inclusion of multiple sites introduces a
source of variability that complicates the analysis. Patients
within a site tend to be more similar to one another than to
patients at other sites, owing to shared clinical practices,
patient populations, and local standards of care. This
within-site correlation, or clustering, violates the independence
assumption underlying many standard statistical methods
[@Localio2001adjustments].

## Regulatory context

Both international and national regulatory bodies have addressed
the role of covariates, including site, in the analysis of
clinical trials. The ICH E9 guideline states that 'factors on
which randomisation has been stratified should be accounted for
later in the analysis' [@ICHE9guideline]. When randomization is
stratified by site, this implies that site should appear in the
analysis model. The United States Food and Drug Administration
(FDA) 2023 guidance on covariate adjustment reinforces this
position, recommending that stratification variables be included
in the primary analysis model [@FDA2023covariate]. The European
Medicines Agency (EMA) similarly advises that stratification
factors be incorporated unless stratification was performed for
purely administrative reasons [@EMA2015covariate].

## The case for adjusting for site

The primary statistical argument for including site in the
analysis model is that it can reduce residual variance, thereby
increasing the precision of treatment effect estimates and
improving power [@Kahan2014rewards]. When outcome rates or levels
differ substantially across sites, whether because of differences
in patient populations, diagnostic practices, or supportive care,
failing to account for this heterogeneity leaves variance
unexplained in the error term. Adjustment for site absorbs this
between-site variability and can yield narrower confidence
intervals for the treatment effect [@Kahan2013continuous;
@edgarIncludingRandomCentre2021].

A further argument concerns the validity of inference when
randomization has been stratified by site. @kahanImproperAnalysisTrials2012
demonstrated that stratified randomization induces correlation
between treatment groups within strata. If the analysis ignores
this structure, standard errors are biased upward, confidence
intervals are too wide, and power is reduced.
@linschotenIncorrectlyAnalysingStratified2022 provided a
clinical example in which failure to adjust for stratification
factors led to a false negative conclusion.

## Fixed versus random effects

A central debate concerns whether site should enter the model as
a fixed or random effect. Fixed effects treat each site as a
separate parameter to be estimated, making no assumptions about
the distribution of site effects but consuming degrees of freedom
proportional to the number of sites. Random effects assume site
effects are drawn from a normal distribution, estimating only the
variance of that distribution and thereby requiring fewer
parameters [@Falissard1996center].

@Kahan2013continuous conducted a simulation study comparing fixed
and random centre effects for continuous outcomes across 378
scenarios. Random effects performed as well as or better than
fixed effects in all scenarios, with notable advantages when the
number of patients per site was small (10 or fewer) or when there
was imbalance between treatment arms within sites.
@Kahan2014binary extended these findings to binary outcomes,
comparing fixed effects, random effects, generalized estimating
equations (GEEs), and Mantel-Haenszel methods.

The practical relevance of this comparison is underscored by
@Kahan2015fewevents, who reviewed 206 multicenter trials in four
major medical journals and found that the median number of events
per centre-treatment arm combination was just 3, with 50% of
trials having fewer than 3 events per centre-treatment
combination. Fixed-effects models require substantial data per
site to perform well; in the majority of real trials, this
condition is not met [@kimAnalysisMulticenterClinical2020].

## Treatment-by-site interaction

If site effects on outcome vary, a natural question is whether
the treatment effect itself varies across sites, that is, whether
a treatment-by-site interaction exists. A qualitative interaction,
where the direction of the treatment effect reverses at some
sites, has major clinical implications: it suggests that the
treatment may be harmful for some patient populations
[@GailSimon1985qualitative]. However, tests for treatment-by-site
interaction are typically underpowered because individual sites
rarely enroll enough patients to estimate site-specific treatment
effects precisely [@edgarIncludingRandomCentre2021].

@Agresti2000strategies discussed strategies for handling
treatment-by-centre interaction with binary outcomes, noting that
random effects models smooth extreme site-specific estimates and
provide more stable inference than fixed effects when some sites
contribute sparse data. @Islam2024centerlevel reviewed recent
practice and confirmed that the choice of method for handling
centre effects can materially affect trial conclusions.

## Heterogeneous site sizes

Real multicenter trials rarely have equal enrollment across
sites. Some sites may contribute hundreds of patients while
others contribute fewer than ten. This heterogeneity in site
size interacts with the choice of analytic method.
Fixed-effects models can lead to patient exclusion when sites
have observations in only one treatment arm (a common occurrence
with small sites), whereas random-effects models borrow strength
across sites and avoid this problem
[@Kahan2015fewevents; @kimAnalysisMulticenterClinical2020].
@Localio2001adjustments provided an overview of the analytic
challenges introduced by clustering, confounding, and effect
modification in multicenter studies, emphasizing that the
correlation of outcomes within sites must be addressed to obtain
valid inference.

## Present study

Despite the extensive methodological literature, uncertainty
persists about when and how to adjust for site in practice.
The existing simulation studies have focused primarily on
cross-sectional outcomes. In many clinical trials, however,
outcomes are measured longitudinally over multiple visits,
introducing an additional layer of within-subject correlation that
interacts with the between-site structure. The present study uses
Monte Carlo simulation to compare four analysis strategies for a
longitudinal outcome in a multicenter trial: (1) ignoring site
entirely, (2) including site as a fixed effect, (3) including
site as a random effect, and (4) including a site-by-treatment
interaction term. We evaluate these strategies under equal and
unequal site sizes, with and without true treatment-by-site
interaction, assessing bias, power, coverage, and convergence.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Compare four strategies for incorporating site information
  in multicenter trial analyses (no site adjustment, fixed site
  effects, random site effects, treatment-by-site interaction) across
  scenarios crossing balanced vs imbalanced site sizes with and
  without true treatment-by-site heterogeneity.
- **Data-generating mechanism.** Detailed in the next subsection.
- **Estimand.** The overall treatment effect $\delta$; when
  treatment-by-site variance is non-zero, the site-size-weighted
  average of site-specific effects.
- **Methods.** Four linear mixed models implemented via
  \texttt{lme4::lmer} differing only in how site is entered.
- **Performance measures.** Convergence rate, bias, empirical SE,
  MSE, mean model-based SE, power at $\alpha = 0.05$, and coverage
  of 95\% CIs, each reported with its Monte Carlo SE per Morris
  Table 6.

## Data generating model

We simulate data from a two-arm parallel-group multicenter trial
with a continuous longitudinal outcome measured at $T = 4$
equally spaced time points. The model for the outcome $Y_{ijk}$
of subject $j$ at site $i$ and time $k$ is:

$$
Y_{ijk} = \alpha_i + (\beta + \gamma_i) \cdot \text{Trt}_{ij}
  + \delta \cdot t_k + u_j + \epsilon_{ijk}
$$

where $\alpha_i \sim N(0, \sigma^2_s)$ is the random site
intercept, $\beta$ is the true treatment effect,
$\gamma_i \sim N(0, \sigma^2_{s \times t})$ is the random
treatment-by-site interaction, $\delta = 0.05$ is a linear time
slope, $u_j \sim N(0, 0.25)$ is a subject-level random
intercept, and $\epsilon_{ijk} \sim N(0, \sigma^2_e)$ is
residual error.

## Simulation scenarios

We examine four scenarios defined by crossing two factors:

\begin{itemize}
  \item \textbf{Site size}: equal (20 patients per site) versus
    unequal (exponentially distributed with mean 20,
    minimum 4)
  \item \textbf{Treatment-by-site interaction}: absent
    ($\sigma^2_{s \times t} = 0$) versus present
    ($\sigma^2_{s \times t} = 0.15$)
\end{itemize}

In all scenarios, we set $K = 20$ sites, $\beta = 0.30$,
$\sigma^2_s = 0.25$, and $\sigma^2_e = 1.0$. Within each
site, subjects are randomized 1:1 to treatment and control.

## Analysis models

For each simulated dataset, we fit four linear mixed models
using the \texttt{lme4} package [@R-lme4]:

\begin{enumerate}
  \item \textbf{No site}: \texttt{y \~{} trt + time + (1|subj)}
  \item \textbf{Site fixed}:
    \texttt{y \~{} trt + time + site + (1|subj)}
  \item \textbf{Site random}:
    \texttt{y \~{} trt + time + (1|site) + (1|subj)}
  \item \textbf{Site $\times$ treatment}:
    \texttt{y \~{} trt*site + time + (1|subj)}
\end{enumerate}

## Performance metrics

For each model across 1,000 replications, we compute:

- **Bias**: mean estimated treatment effect minus true value
  ($\beta = 0.30$)
- **Empirical SE**: standard deviation of estimates across
  replications
- **Mean model SE**: average of model-based standard errors
- **Power**: proportion of replications with $p < 0.05$
- **Coverage**: proportion of nominal 95\% confidence intervals
  containing the true value
- **Convergence rate**: proportion of replications in which the
  model converged

```{r sim-seed, cache=FALSE}
# Morris, White, and Crowther (2019) §4.1: pin RNGkind and set the
# seed ONCE at the start of the program. `sim_site_covariate()` no
# longer accepts a `seed` argument; each scenario consumes state from
# this single stream in sequence. Differences between scenarios then
# reflect DGM differences, not distinct RNG starting points.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260312)
```

For coverage MCSE $\leq 0.7$ pp at 0.95, $n_{\mathrm{sim}} \geq 970$;
$n_{\mathrm{sim}} = 1000$ gives coverage MCSE $\approx 0.69$ pp.

```{r run-sim-equal-nointeract, cache=TRUE}
res_eq_no <- sim_site_covariate(
  n_sim = 1000, n_sites = 20,
  site_size = "equal", true_trt = 0.30,
  site_sd = 0.50, trt_site_sd = 0.0
)
```

```{r run-sim-unequal-nointeract, cache=TRUE}
res_uneq_no <- sim_site_covariate(
  n_sim = 1000, n_sites = 20,
  site_size = "unequal", true_trt = 0.30,
  site_sd = 0.50, trt_site_sd = 0.0
)
```

```{r run-sim-equal-interact, cache=TRUE}
res_eq_int <- sim_site_covariate(
  n_sim = 1000, n_sites = 20,
  site_size = "equal", true_trt = 0.30,
  site_sd = 0.50, trt_site_sd = 0.15
)
```

```{r run-sim-unequal-interact, cache=TRUE}
res_uneq_int <- sim_site_covariate(
  n_sim = 1000, n_sites = 20,
  site_size = "unequal", true_trt = 0.30,
  site_sd = 0.50, trt_site_sd = 0.15
)
```

# Results

## Headline findings

```{r headline-11, include=FALSE}
summ_eq_no   <- summarise_sim(res_eq_no,   true_trt = 0.30)
summ_uneq_no <- summarise_sim(res_uneq_no, true_trt = 0.30)
summ_eq_int  <- summarise_sim(res_eq_int,  true_trt = 0.30)
summ_uneq_int<- summarise_sim(res_uneq_int,true_trt = 0.30)
all_summ <- dplyr::bind_rows(
  dplyr::mutate(summ_eq_no, scenario = "Equal, no interaction"),
  dplyr::mutate(summ_uneq_no, scenario = "Unequal, no interaction"),
  dplyr::mutate(summ_eq_int, scenario = "Equal, with interaction"),
  dplyr::mutate(summ_uneq_int, scenario = "Unequal, with interaction")
)
bias_max <- all_summ |>
  dplyr::slice_max(abs(bias), n = 1, with_ties = FALSE)
cov_range <- range(all_summ$coverage, na.rm = TRUE)
pow_by_model <- all_summ |>
  dplyr::group_by(model) |>
  dplyr::summarise(pow_mean = mean(power, na.rm = TRUE))
```

Across four scenarios crossing balanced vs imbalanced
site sizes with and without true
treatment-by-site heterogeneity, all four analysis models
produced absolute bias below
`r sprintf('%.3f', abs(bias_max$bias))` (worst case:
`r bias_max$model` under "`r bias_max$scenario`", MCSE
`r sprintf('%.3f', bias_max$mcse_bias)`). Coverage of 95\%
confidence intervals across all combinations fell in
[`r sprintf('%.3f', cov_range[1])`,
`r sprintf('%.3f', cov_range[2])`]. Mean power by model
(averaged across scenarios) was:
`r paste(sprintf('%s = %.2f', pow_by_model$model,
pow_by_model$pow_mean), collapse = ', ')`. All performance
measures are reported with their Morris Table 6
Monte Carlo SEs in the per-scenario tables below.

## Scenario 1: Equal site sizes, no interaction

```{r tab-equal-no}
tab1 <- summarise_sim(res_eq_no, true_trt = 0.30)
tab1 |>
  select(model, bias, empirical_se, mean_model_se,
         power, coverage, convergence_rate, n_converged) |>
  mutate(
    model = c("No site", "Site (fixed)",
              "Site (random)", "Site x Trt")
  ) |>
  kable(
    format = "latex", booktabs = TRUE,
    caption = paste(
      "Performance of four analysis models: equal",
      "site sizes, no treatment-by-site interaction."
    ),
    digits = 3, escape = FALSE,
    col.names = c(
      "Model", "Bias", "Emp. SE",
      "Model SE", "Power", "Coverage",
      "Conv.", "N conv."
    )
  ) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

## Scenario 2: Unequal site sizes, no interaction

```{r tab-unequal-no}
tab2 <- summarise_sim(res_uneq_no, true_trt = 0.30)
tab2 |>
  select(model, bias, empirical_se, mean_model_se,
         power, coverage, convergence_rate, n_converged) |>
  mutate(
    model = c("No site", "Site (fixed)",
              "Site (random)", "Site x Trt")
  ) |>
  kable(
    format = "latex", booktabs = TRUE,
    caption = paste(
      "Performance of four analysis models: unequal",
      "site sizes, no treatment-by-site interaction."
    ),
    digits = 3, escape = FALSE,
    col.names = c(
      "Model", "Bias", "Emp. SE",
      "Model SE", "Power", "Coverage",
      "Conv.", "N conv."
    )
  ) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

## Scenario 3: Equal site sizes, with interaction

```{r tab-equal-int}
tab3 <- summarise_sim(res_eq_int, true_trt = 0.30)
tab3 |>
  select(model, bias, empirical_se, mean_model_se,
         power, coverage, convergence_rate, n_converged) |>
  mutate(
    model = c("No site", "Site (fixed)",
              "Site (random)", "Site x Trt")
  ) |>
  kable(
    format = "latex", booktabs = TRUE,
    caption = paste(
      "Performance of four analysis models: equal",
      "site sizes, with treatment-by-site interaction."
    ),
    digits = 3, escape = FALSE,
    col.names = c(
      "Model", "Bias", "Emp. SE",
      "Model SE", "Power", "Coverage",
      "Conv.", "N conv."
    )
  ) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

## Scenario 4: Unequal site sizes, with interaction

```{r tab-unequal-int}
tab4 <- summarise_sim(res_uneq_int, true_trt = 0.30)
tab4 |>
  select(model, bias, empirical_se, mean_model_se,
         power, coverage, convergence_rate, n_converged) |>
  mutate(
    model = c("No site", "Site (fixed)",
              "Site (random)", "Site x Trt")
  ) |>
  kable(
    format = "latex", booktabs = TRUE,
    caption = paste(
      "Performance of four analysis models: unequal",
      "site sizes, with treatment-by-site interaction."
    ),
    digits = 3, escape = FALSE,
    col.names = c(
      "Model", "Bias", "Emp. SE",
      "Model SE", "Power", "Coverage",
      "Conv.", "N conv."
    )
  ) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

## Comparison of bias and power across scenarios

```{r fig-bias, fig.width=5.5, fig.height=3.5, fig.cap="Bias of the treatment effect estimate across four models and four simulation scenarios."}
all_results <- bind_rows(
  res_eq_no |> mutate(
    scenario = "Equal, no interaction"),
  res_uneq_no |> mutate(
    scenario = "Unequal, no interaction"),
  res_eq_int |> mutate(
    scenario = "Equal, with interaction"),
  res_uneq_int |> mutate(
    scenario = "Unequal, with interaction")
)

all_results |>
  filter(converged) |>
  mutate(
    model = factor(
      model,
      levels = c("no_site", "site_fixed",
                  "site_random", "site_x_trt"),
      labels = c("No site", "Fixed",
                 "Random", "Site x Trt")
    )
  ) |>
  ggplot(aes(x = model, y = estimate - 0.30)) +
  geom_hline(
    yintercept = 0, linetype = "dashed",
    colour = "grey50"
  ) +
  geom_boxplot(
    fill = "grey85", outlier.size = 0.5
  ) +
  facet_wrap(~scenario, nrow = 1) +
  labs(
    x = "Analysis model",
    y = "Bias (estimate - 0.30)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1, size = 7
    )
  )
```

```{r fig-power, fig.width=5.5, fig.height=3.5, fig.cap="Power to detect the treatment effect across four models and four simulation scenarios."}
power_dat <- all_results |>
  filter(converged) |>
  mutate(
    model = factor(
      model,
      levels = c("no_site", "site_fixed",
                  "site_random", "site_x_trt"),
      labels = c("No site", "Fixed",
                 "Random", "Site x Trt")
    )
  ) |>
  group_by(model, scenario) |>
  summarise(
    power = mean(pvalue < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

power_dat |>
  ggplot(aes(
    x = model, y = power,
    shape = scenario, group = scenario
  )) +
  geom_point(size = 2.5) +
  geom_line(aes(linetype = scenario)) +
  scale_shape_manual(
    values = c(16, 17, 1, 2)
  ) +
  scale_linetype_manual(
    values = c("solid", "dashed",
               "dotted", "dotdash")
  ) +
  labs(
    x = "Analysis model", y = "Power",
    shape = "Scenario", linetype = "Scenario"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```

```{r fig-coverage, fig.width=5.5, fig.height=3.5, fig.cap="Coverage of nominal 95 percent confidence intervals across four models and four simulation scenarios. The dashed line indicates the nominal 95 percent level."}
cov_dat <- all_results |>
  filter(converged) |>
  mutate(
    covered = (estimate - 1.96 * se <= 0.30) &
      (estimate + 1.96 * se >= 0.30),
    model = factor(
      model,
      levels = c("no_site", "site_fixed",
                  "site_random", "site_x_trt"),
      labels = c("No site", "Fixed",
                 "Random", "Site x Trt")
    )
  ) |>
  group_by(model, scenario) |>
  summarise(
    coverage = mean(covered, na.rm = TRUE),
    .groups = "drop"
  )

cov_dat |>
  ggplot(aes(
    x = model, y = coverage,
    shape = scenario, group = scenario
  )) +
  geom_hline(
    yintercept = 0.95, linetype = "dashed",
    colour = "grey50"
  ) +
  geom_point(size = 2.5) +
  geom_line(aes(linetype = scenario)) +
  scale_shape_manual(
    values = c(16, 17, 1, 2)
  ) +
  scale_linetype_manual(
    values = c("solid", "dashed",
               "dotted", "dotdash")
  ) +
  coord_cartesian(ylim = c(0.85, 1.0)) +
  labs(
    x = "Analysis model", y = "Coverage",
    shape = "Scenario", linetype = "Scenario"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```

# Discussion

The simulation results illuminate several practical
considerations for the inclusion of site in the analysis of
multicenter trials with longitudinal outcomes.

First, consistent with the findings of @Kahan2013continuous
and @edgarIncludingRandomCentre2021, including site as a
random effect generally performs as well as or better than
fixed effects across all scenarios examined. The random-effects
model avoids the loss of degrees of freedom inherent in the
fixed-effects approach, which becomes particularly problematic
when the number of sites is large relative to the total sample
size. This advantage is amplified when site sizes are unequal,
as fixed-effects models can produce unstable estimates when
some sites contribute very few subjects per treatment arm
[@Kahan2015fewevents; @kimAnalysisMulticenterClinical2020].

Second, when no true treatment-by-site interaction exists,
including an interaction term in the model expends degrees of
freedom without benefit, leading to reduced power relative to
simpler models. This finding supports the common recommendation
to include the interaction term only when there is a strong
a priori reason to expect treatment effect heterogeneity
across sites [@GailSimon1985qualitative;
@Agresti2000strategies].

Third, the regulatory guidance from ICH E9, the FDA, and the
EMA all converge on the recommendation that stratification
variables should be included in the analysis model
[@ICHE9guideline; @FDA2023covariate; @EMA2015covariate]. Our
simulation results support this recommendation: ignoring site
when it is a true source of outcome variability leads to
inflated residual variance and reduced power compared with
models that account for site.

Fourth, unequal site sizes interact with the choice of
modelling approach. The random-effects model naturally handles
imbalanced designs by borrowing information across sites,
whereas the fixed-effects model treats each site independently
and can perform poorly when some sites contribute sparse data
[@Localio2001adjustments; @Agresti2000strategies].

Several limitations should be noted. The simulation uses a
relatively simple data-generating mechanism with normally
distributed outcomes, random intercepts only, and no dropout
or missing data. Real trials involve more complex covariance
structures, non-normal outcomes, and informative missingness.
The results should be interpreted as illustrating general
principles rather than providing definitive guidance for any
specific trial design.

# Future research

Several directions merit further investigation:

1. **Non-normal outcomes.** Extending the simulation to binary
   and time-to-event endpoints would complement the present
   results for continuous outcomes and address the findings
   of @Kahan2014binary and
   @kimAnalysisMulticenterClinical2020 regarding sparse
   events per site.

2. **Complex covariance structures.** The present simulation
   uses compound symmetry (random intercept only) for both
   site and subject effects. Examining unstructured or
   autoregressive within-subject correlations, and allowing
   random slopes at the site level, would better approximate
   real longitudinal trial data.

3. **Informative site sizes.** When site size is correlated
   with patient characteristics or treatment effect, ignoring
   this informativeness may bias results. Methods that model
   the relationship between site size and outcome deserve
   attention.

4. **Adaptive designs.** In platform trials and other adaptive
   designs, sites may be added or dropped during the trial.
   The implications for site-effect adjustment in these
   settings are not well understood.

5. **Software tools.** Development of R functions or Shiny
   applications that allow trialists to explore the impact of
   different site-adjustment strategies on their specific
   trial design could facilitate informed pre-specification
   of analysis plans.

6. **Bayesian approaches.** Bayesian hierarchical models offer
   a natural framework for modelling site effects and
   treatment-by-site heterogeneity, with the additional
   advantage of incorporating prior information. Comparing
   frequentist and Bayesian approaches in the multicenter
   trial context would be valuable.

7. **Empirical review.** A systematic review of how site
   effects are handled in recently published trials in major
   medical journals (NEJM, JAMA, Lancet) would quantify
   current practice and identify gaps between methodological
   recommendations and applied work, extending the review by
   @Kahan2015fewevents.

# References

## Morris et al. (2019) ADEMP Compliance

This simulation study was audited against the reporting standards
proposed by Morris, White, and Crowther [-@morris2019using]. The full
audit is at `docs/morris-audit-2026-04-17.md`.

**Verdict:** Partially compliant.

**Key gaps identified:**

- `set.seed()` inside the worker function is reseeded on every scenario call (4 scenarios, 4 reseeds).
- No Monte Carlo SE on any metric (`summarise_sim()` at lines 166-184).
- `n_sim = 1000` hardcoded without a target-MCSE derivation.

A remediation plan is documented in the audit file and will be
executed in a subsequent revision.
